---
title: "sta309-Midterm_RM"
author: "Ryan McCollum"
date: "2023-04-05"
output: html_document
---

# Evaluating the Effects of Economic Disparities on State Math Tests 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("cowplot")
library(tidyverse)
library(ggrepel)
library(ggthemes)
library(patchwork)
library(mplot)
library(ggplot2)
library(scales)
library(cowplot)
```

### Loading in files
```{r, cache=TRUE, warning=FALSE}
all_SD_math_20_21 <- read.csv("math-achievement-lea-sy2020-21.csv") %>% 
  group_by(LEAID) %>%
  filter(GRADE == "00") %>%
  filter(CATEGORY == "ALL" | CATEGORY == "ECD") %>%
  select(-SCHOOL_YEAR, -FIPST, -ST_LEAID, -SUBJECT, -DATE_CUR) %>%
  mutate(PCTPROF = as.numeric(PCTPROF)) %>%
  na.omit()
```


## Economic Disadvantages by State
```{r, warning=FALSE}

## Map data set
US_states <- map_data("state") %>%
  mutate(region=str_to_upper(region))

## Data set with all valid entries for each school
US_total_district_math <- all_SD_math_20_21 %>%
  na.omit(all_SD_math_20_21) %>%
  group_by(LEANM) %>%
  filter(CATEGORY =="ALL",
         GRADE == "00") %>%
  mutate(TotalPctProf = PCTPROF,
         all_valid = NUMVALID) %>%
  select(LEAID, all_valid, TotalPctProf, STNAM)

## Data set with only ECD entries
US_ECD_math <- all_SD_math_20_21 %>% 
  filter(CATEGORY =="ECD",
         GRADE == "00")

## Merging ECD and ALL data sets by school district
US_compare_ECD <- merge(US_ECD_math, US_total_district_math,
                               by.x="LEAID", by.y="LEAID") %>%
  group_by(STNAM.x) %>%
  summarize(sum_valid_ECD = sum(NUMVALID),
            sum_NUMVALID = sum(all_valid),
            avg_mathProf = mean(PCTPROF)) %>%
  mutate(PCT_ECD = (sum_valid_ECD/sum_NUMVALID),
                    STNAM = STNAM.x,
         SCHOOL_YEAR = "2020-2021") %>%
  filter(STNAM.x != "ALASKA") %>%
  filter(STNAM.x != "BUREAU OF INDIAN EDUCATION") %>% 
  select(SCHOOL_YEAR, STNAM, avg_mathProf, PCT_ECD) 

## Merging compare_ECD data set with Map
US_ECD_states <- US_compare_ECD %>%
  left_join(US_states, by=c("STNAM"="region"))


## Plotting data 
USA_ECD_math_plot <- ggplot(US_ECD_states, aes(x=long, y=lat, 
                                                     group=group, fill=PCT_ECD)) +
  geom_polygon(color="grey30") + 
  geom_polygon(data=filter(US_states, region=="WASHINGTON"), aes(x=long, y=lat, group=group), fill="white", color="grey30") +
  coord_map("conic", lat=48) +
  scale_fill_gradient2(low="white", mid="goldenrod1", high="firebrick",
                       limits=c(0, 1), midpoint=0.5) +
  annotate(geom="text", x=-120, y=47.5, label="N/A", size=1) +
  theme_map() +
  theme(legend.position = "none",
        plot.title=element_text(hjust=0, size=11.5),
        plot.subtitle=element_text(hjust=0, size=8)) +
  labs(title = "2020-2021 Proportion of Student Entries that are \"Economically Disadvantaged\"",
       subtitle="The number of valid math test entries that were labeled as taken by an \"Economically Disadvantaged\"\n(ECD)  were divided by the total number of valid math test entires from each state to get the\npercentage of entries in each state taken by ECD students. \nNOTES: Washington did not have and school districts with valid entries\n             All of Illinois' entries were labeled as \"ECD\"")
USA_ECD_math_plot
```


```{r, warning=FALSE}
## Ohio Map data set
ohio_map <- map_data("county") %>%
  filter(region=="ohio")
## I edited the CSV in excel to add the generalRegion variable
##    to each Ohio School district
## write.csv(ohio_map, "ohio_map_counties.csv")

  
## All valid entries in each Ohio school district
ohio_total_district_math <- all_SD_math_20_21 %>%
  na.omit(all_SD_math_20_21) %>%
  group_by(LEANM) %>%
  filter(STNAM=="OHIO",
         CATEGORY =="ALL",
         GRADE == "00") %>%
  mutate(all_valid = NUMVALID) %>%
  mutate(District_ID = LEAID - 3900000) %>%
  select(LEANM, all_valid, PCTPROF, CATEGORY, STNAM, GRADE, LEAID, District_ID)

## All ECD entries in each Ohio school district
ohio_ECD_math <- all_SD_math_20_21 %>% 
  filter(STNAM=="OHIO",
         CATEGORY =="ECD",
         GRADE == "00") %>%
  mutate(District_ID = LEAID - 3900000) %>%
  select(LEAID, LEANM, NUMVALID, CATEGORY, District_ID)



## Merging ECD and ALL data sets by school district
##    Creating new variable that determines if more than 50% 
##    of all entries were labeled as ECD for each Ohio school district
ohio_compare_ECD_math <- merge(ohio_ECD_math, ohio_total_district_math,
                               by.x="LEAID", by.y="LEAID") %>%
  mutate(PCT_ECD = NUMVALID / all_valid,
         CATEGORY = case_when((PCT_ECD >= 0.50)~ "Majority Economically Disadvantaged",
                              PCT_ECD < 0.50 ~ "Majority Economically Advantaged"),
         SCHOOL_YEAR = 2021,
         District_ID = LEAID - 3900000) %>%
  na.omit() %>%
  select(SCHOOL_YEAR, STNAM, GRADE, CATEGORY, PCTPROF, LEAID, PCT_ECD, District_ID, NUMVALID, all_valid)


## Making box plot
ohio_box <- ggplot(data=ohio_compare_ECD_math) +
  geom_boxplot(aes(x=CATEGORY, y=PCTPROF), 
               color=c("forestgreen", "firebrick"), size=0.75) +
  geom_boxplot(data=ohio_total_district_math, aes(x="All Schools Combined", 
                                                  y=PCTPROF), size=1) +
  scale_y_continuous(limits=c(0, 100),
                     breaks=c(0, 25, 50, 75, 100),
                     minor_breaks = NULL) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        plot.title=element_text(hjust=0, size=11, face="bold"),
        plot.subtitle=element_text(hjust=0, size=8, color="grey30"),
        plot.caption=element_text(size=8, hjust=1, face="bold")) +
  labs(title="Proficiency of Economically Advantaged vs. Disadvantaged Schools in Ohio",
       subtitle= "School districts were labels as \"Majority Economically Disadvantaged\" if more than 50%\nof their valid math test entries were labeled as ECD",
       caption="SOURCES: US Department of Education\nUS Census")
ohio_box
```





### Comparing different categories
```{r}
## Importing list of all Ohio school districts and their County
sd_by_county <- read.csv("sdlist_2-21.csv") %>%
  mutate(District_ID = as.numeric(District_ID))

## Importing my version of Ohio map data
ohio_map_regions <- read.csv("ohio_map_counties.csv") %>%
  mutate(County = subregion)

## Merging Ohio ECD data with School District County data
ohio_sd_county_ECD <- merge(sd_by_county, ohio_compare_ECD_math,
                            by.x="District_ID", by.y="District_ID") %>%
  mutate(County = str_to_lower(County)) %>%
  mutate(County = str_remove(County, " county"))

## Getting percent of ECD for each Ohio County
ohio_ECD_by_county <- ohio_sd_county_ECD %>%
  group_by(County) %>%
  summarize(Total = sum(all_valid),
            Total_ECD = sum(NUMVALID),
            PCT_ECD_County = (Total_ECD / Total)) 

## Adding regions
ohio_ECD_regions <- merge(ohio_ECD_by_county, ohio_map_regions,
                            by="County")
  
  
## Adding map to ECD data
ohio_ECD_map <- ohio_ECD_by_county %>%
  left_join(ohio_map_regions, by="County")


## Plot
Ohio_ECD_math_plot <- ggplot(ohio_ECD_map, 
                             aes(x=long, y=lat,group=County, fill=PCT_ECD_County)) +
  geom_polygon(color="grey30") + 
  geom_polygon(data=filter(ohio_map_regions, County=="fulton"),
               aes(x=long, y=lat, group=County), fill="white", color="grey30") +
  coord_map() +
  scale_fill_gradient2(low="white", mid="goldenrod1", high="firebrick",
                       limits=c(0, 1), midpoint=0.5) +
  annotate(geom="text", x=-84.15, y=41.65, label="N/A", size=1) +
  theme_map() +
  theme(legend.position = "bottom",
        legend.key.width=unit(1.2, "cm"),
        #legend.title = element_text(color = "gray30", size = 10),
        legend.title = element_blank(),
        plot.subtitle = element_text(color="gray30", size=8, hjust=0),
        plot.caption.position = "plot",
        plot.title=element_text(size=8, face="bold")) +
  labs(title="State of Ohio", 
       subtitle="NOTES: Fulton County did not have and school districts with valid entries.\nNot all schools are included, some do not have any ECD entries")
Ohio_ECD_math_plot
```



## Making data set and Model for Comparing ECD and PROFPCT Across School Districts and Regions
```{r}
## Grabbing data for all Ohio schools with ECD entries
ohio_all_valid_schools <- ohio_sd_county_ECD %>%
  left_join(ohio_map_regions, by="County") %>%
  mutate(Entries = all_valid) %>%
  select(District_Name, County, CATEGORY, PCTPROF, PCT_ECD, Entries, generalRegion) %>%
  unique() %>%
  na.omit()

## Grabbing my high school
my_high_school <- ohio_all_valid_schools %>%
  filter(District_Name == "Three Rivers Local School District") %>%
  mutate(generalRegion = "My High School")

## Used to change color of legend
for_legend <- ohio_all_valid_schools %>%
  filter(District_Name == "Three ") %>%
  mutate(generalRegion = "none")

## Finding the best fit model
ohio.fit_lin <- lm(data=ohio_all_valid_schools, PCTPROF~PCT_ECD)
#summary(ohio.fit_lin)
ohio.fit_quad <- lm(PCTPROF~PCT_ECD + I(PCT_ECD^2), data=ohio_all_valid_schools)
#summary(ohio.fit_quad)
ohio.fit_cubic <- lm(PCTPROF~PCT_ECD + I(PCT_ECD^3), data=ohio_all_valid_schools)
#summary(ohio.fit_cubic)

fake_data <- data.frame(
  PCT_ECD = seq(0.0, 1, 0.02)
)
  
fake_data <- fake_data %>%
  mutate(PCTPROFT_fine = predict(ohio.fit_cubic, newdata=fake_data, type="response"))
```
### Comparing Models

ADJR2(ohio.fit_lin) = 0.5661  

ADJR2(ohio.fit_quad) = *0.6047*
AIC(ohio.fit_quad)  = *3224.412*
BIC(ohio.fit_quad)  = *3240.713*


ADJR2(ohio.fit_cubic) = *0.6064* 
AIC(ohio.fit_cubic) = *3222.618*
BIC(ohio.fit_cubic) = *3238.92*

**Cubic model seems to be the best by just a little**


## Comparing all Ohio Schools with valid ECD entries
```{r, warning=FALSE}
## Plot
ohio_regions_schools <- 
  ggplot() +
  geom_point(data=ohio_all_valid_schools, aes(x=PCT_ECD, y=PCTPROF, group=District_Name,  
                                              color=generalRegion, size=Entries)) +
  geom_point(data=my_high_school, aes(x=PCT_ECD, y=PCTPROF, group=District_Name,
                                      size=Entries), color="yellow1") +
  geom_point(data=for_legend, aes(x=PCT_ECD, y=PCTPROF, group=District_Name,
                                      size=Entries), color="grey10") +
  geom_line(data=fake_data, aes(x=PCT_ECD, y=PCTPROFT_fine),
            size=1.25, color="navy", alpha=0.6) +
  geom_smooth(method='lm', formula= y~x) +
  theme_minimal() +
  scale_x_continuous(breaks = c(0.25, 0.5, 0.75, 1),
                     labels=c("25", "50", "75", "100"),
                     minor_break = NULL) +
  scale_y_continuous(breaks = c(25, 50, 75, 100),
                     minor_break = NULL) +
  xlab("% Economically Disadvantaged") +
  ylab("% at or Above Proficiency in Math") +
  annotate(geom="text", label="Three Rivers Local\nSchool District\n(The school I attended)", 
           x=.20, y=26, color="goldenrod1", size=3, fontface="bold") +
  annotate(geom="segment", x=0.21, xend=0.3025, y=34.5, yend=57.5 ) +
  theme(legend.position = "right",
        legend.box = "vertical",
        legend.title = element_text("Number of valid entries", size=6),
        axis.title.x=element_text(size=7, color="grey40"),
        axis.title.y=element_text(size=7, color="grey40"),
        plot.title=element_text(size=11, face="bold"),
        legend.key.width=unit(0.01, "cm"),
        legend.key.size=unit(0.05, "cm"),
        plot.subtitle=element_text(size=8, color="grey30")) +
  guides(color="none") +
  labs(title="Comparing All Ohio Schools with at Least 1 ECD Entry",
       subtitle="Graph compares each schools ECD to their Percent Proficieny\nas well as their counties region in Ohio and the size of the school") +
  annotate("text", x=0.05, y=88, label="Central", color="#F8766D",
           size=4) +
  annotate("text", x=0.615, y=61, label="SouthEast", color="#00BFC4",
           size=4) + 
  annotate("text", x=0.30, y=83, label="NorthEast", color="#7CAE00",
           size=4) + 
  annotate("text", x=0.81, y=8, label="SouthWest", color="#E76BF3",
           size=4) +
  annotate("text", x=0.43, y=31, label="NorthWest", color="seagreen3",
           size=4) +
  annotate("text", x=.75, y=75, label="Line of best fit from\na cubic model",
           color="navy", size=3, fontface="bold") +
  annotate("segment", x=.76, xend=0.68, y=70, yend=39)
 
ohio_regions_schools
```

## Ohio County Regional Map
```{r, warning=FALSE}
## Creating an Ohio plot based on 5 regions for legend purposes
Ohio_regional <- ggplot(ohio_map_regions, 
                        aes(x=long, y=lat, group=County, fill=generalRegion)) +
  geom_polygon(color="grey30") +
  coord_map() +
  scale_x_continuous(breaks=NULL,
                     minor_breaks = NULL) +
  scale_y_continuous(breaks=NULL,
                     minor_breaks = NULL) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y=element_blank(),
        plot.title=element_text(size=8, face="bold")) +
  annotate("point", x=-84.7502 , y=39.2, color="yellow") +
  labs(title="Ohio Counties\nby Region")
Ohio_regional
```






# DASHBOARD
```{r, warning=FALSE}
## Creating Dashboard
empty_plot <- ggplot() +
  theme_minimal()

top_plot <-
  plot_grid(USA_ECD_math_plot, ohio_regions_schools, ncol=2,
                      rel_widths=c(0.50, 0.50), rel_heights=c(.60, .40))


legend <- 
  plot_grid(empty_plot, Ohio_regional, ncol=2, rel_widths=c(0.10, 0.9))


top_plot_with_legend <-
  plot_grid(top_plot, legend, ncol=2, rel_widths = c(0.90, 0.1))

bottom_plot <- 
  plot_grid(Ohio_ECD_math_plot, ohio_box, ncol=2, 
            rel_widths=c(0.5, 0.5))

FINAL_plot <- 
  plot_grid(top_plot_with_legend, bottom_plot, nrow=2, rel_heights=c(0.5, 0.50))
#FINAL_plot
```






### Saving Image
```{r}
ggsave(filename="sta309_Midterm-Dashboard.png",
       width=16, height=7, dpi=900, bg="white")

```
